## Determination of white wine quality by exploratory data analysis and multiple linear regression model using R


### By Lasantha Rathnayake

In this exploratory data analysis study, white wine Quality data is examined.

The dataset used in this investigation is a part of two datasets (for white wine
and red whine) from the following is the source.

*   P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
    Modeling wine preferences by data mining from physicochemical properties.
    In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
## Installing required packages
# install.packages("ggplot2", dependencies = T) 
# install.packages("knitr", dependencies = T)
# install.packages("dplyr", dependencies = T)
#install.packages('GGally', dependencies = T)

## Load required packages
library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(GGally)

```


### Information about dataset

In the dataset, there are about 4900 records and 12 attributes. Most 
of the attributes are chemical and physical properties. The data were obtained 
experimentally for different wine types.

The dataset contains following attributes.

```{r echo = FALSE, Load_the_Data}
# Load the Data
#getwd()
#list.files()
win_data =  read.csv('wineQualityWhites.csv', sep = ',')
names(win_data)

```
The size of the dataset is as follows.
```{r, echo = FALSE}
## Exclude the unwanted variable
win_data$X = NULL

dim(win_data)
```
In the original source, followings are the descriptions of attributes in the 
dataset.The corresponding units if any for the attributes are also given in 
along with the attribute name.

fixed acidity (g/dm^3):

*   Most acids involved with wine or fixed or nonvolatile 
    (do not evaporate readily).

volatile acidity (g/dm^3):

*   The amount of acetic acid in wine, which at too high of levels can lead to 
    an unpleasant, vinegar taste

citric acid (g/dm^3):

*   Found in small quantities, citric acid can add 'freshness' and flavor 
    to wines

residual sugar (g/dm^3):

*   The amount of sugar remaining after fermentation stops, it's rare to 
    find wines with less than 1 gram/liter and wines with greater than 
    45 grams/liter are considered sweet

chlorides (g/dm^3): 

*   The amount of salt in the wine

free sulfur dioxide (g/dm^3):

*   The free form of SO2 exists in equilibrium between molecular SO2 
    (as a dissolved gas) and bisulfite ion; it prevents microbial 
    growth and the oxidation of wine

total sulfur dioxide (g/dm^3):

*   Amount of free and bound forms of S02; in low concentrations, SO2 is mostly 
    undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 
    becomes evident in the nose and taste of wine

density (g/cm^3): 

*   The density of water is close to that of water depending on the percent 
    alcohol and sugar content

pH: 
*   Describes how acidic or basic a wine is on a scale from 0 (very acidic) to 
    14 (very basic); most wines are between 3-4 on the pH scale

sulfates (g/dm^3):

*   A wine additive which can contribute to sulfur dioxide gas (S02) levels, 
    which acts as an antimicrobial and antioxidant

alcohol (% by volume): 

*   The percent alcohol content of the wine

quality: 

*   Score between 0 and 10



In order to get a general idea of the data, the following snippet of data was 
obtained (only the first six records of the dataset).

```{r, echo = FALSE}
head(win_data)
```

All the attributes in the dataset are numerical. However, The 'quality' 
attribute was factorized, because the it is used to rank the wine quality.


### Univariate Plots Selection and analysis

Lets look at the distribution of each individual variable as histograms. For 
each variable, a summary of the data distribution is given below the plot.

```{r, echo = FALSE}
## Function to create a histogram
create_histogram = function(dat_set, variabl, binwidth = 0.25) {
                    return( ggplot(aes_string(x = variabl), data = dat_set) + 
                            geom_histogram(bins = 35, binwidth = binwidth,
                                           color = 'black', fill = '#F79420') )
}


```


```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, 'fixed.acidity')

```

The data in the fixed.acidity variable is normally distributed with following 
statistics.

```{r, echo = FALSE}
var = win_data$fixed.acidity
summary(var)
```
            
```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "volatile.acidity", 0.05)

```

The data in the volatile.acidity variable is normally distributed with following
statistics.

```{r, echo = FALSE}
var = win_data$volatile.acidity
summary(var)
```

```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "citric.acid", 0.05)

```

The data in the citric.acid variable is normally distributed with following 
statistics.
```{r, echo = FALSE}
var = win_data$citric.acid
summary(var)
```



```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "residual.sugar", 1)

```

The data in the residual.sugar variable is left skewed with following 
statistics.
```{r, echo = FALSE}
var = win_data$residual.sugar
summary(var)
```

When comparing the maximum value residual.sugar content with 3rd quartile, there
is a huge different can be observable. This is probably due to outliers. In the 
following, plot the outliers were removed. 

```{r, echo = FALSE}
## Call function to create a histogram
win_data_x = win_data[win_data$residual.sugar < 30,]

create_histogram(win_data_x, "residual.sugar", 1)

```

After removing the outliers, following is the summary for residual.sugar 
variable. Th removal of outliers give much better visualization for the data 
distribution. 

```{r, echo = FALSE}
var = win_data_x$residual.sugar
summary(var)
```


```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "chlorides", 0.007)

```

The data in the chlorides variable is normally distributed with following 
statistics. The extreme values of chlorides are difficult to determine without
further information.

```{r, echo = FALSE}
var = win_data$chlorides
summary(var)
```


```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "free.sulfur.dioxide", 6)

```

The data in the free.sulfur.dioxide variable is normally distributed with 
following statistics.
```{r, echo = FALSE}
var = win_data$free.sulfur.dioxide
summary(var)
```

In free.sulfur.dioxide also there are some potential outliers when compare 
maximum value with the third quartile. Therefore, the extreme values should be 
removed.

```{r, echo = FALSE}
win_data_y = win_data[win_data$free.sulfur.dioxide < 200,]
## Call function to create a histogram
create_histogram(win_data_y , "free.sulfur.dioxide", 6)

```
After removing extreme values of free.sulfur.dioxide, following summary 
statistics can be observable.

```{r, echo = FALSE}
var = win_data_y$free.sulfur.dioxide
summary(var)
```


```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "total.sulfur.dioxide", 8)

```
The data in the total.sulfur.dioxide variable is normally distributed with 
following statistics.

```{r, echo = FALSE}
var = win_data$total.sulfur.dioxide
summary(var)
```



```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "density", 0.001)

```


The data in the density variable is normally distributed with following 
statistics.

```{r, echo = FALSE}
var = win_data$density
summary(var)
```

Generally, in standard conditions, the density of water is 1.00 g/cm^3. 
Therefore, the density of a solution in which the water is the solvent, should 
be slightly less than 1.00 g/cm^3. In the above plot we can observe at least 
two density values greater than 1.00 g/cm^3. Therefore, we can remove those 
outliers from the dataset to clean the dataset. Following plot was generated 
without the density outliers. 


```{r, echo = FALSE}
win_data_z = win_data[win_data$density < 1.00,]

density_no_outliers = win_data_x$density

## Call function to create a histogram
create_histogram(win_data_z , "density", 0.001)

```

After removing extreme values of density variable, following summary 
statistics can be observable.

```{r, echo = FALSE}
var = win_data_x$density
summary(var)
```

```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "pH", 0.05)

```

The data in the pH variable is normally distributed with following statistics.
Compare to other variables, the pH values show better Gaussian distribution.

```{r, echo = FALSE}
var = win_data$pH
summary(var)
```


```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "sulphates", 0.02)

```

The data in the sulfates variable is normally distributed with following statistics.
```{r, echo = FALSE}
var = win_data$sulphates
summary(var)
```


```{r, echo = FALSE}
## Call function to create a histogram
create_histogram(win_data, "alcohol", 0.25)

```

The data in the citric.acid variable is somewhat normally distributed with 
following statistics.

```{r, echo = FALSE}
var = win_data$alcohol
summary(var)
```


```{r, echo = FALSE}
## Call function to create a histogram
win_data[, 'quality'] = as.factor(win_data[, 'quality'])
ggplot(aes(x = quality), data = win_data) + 
        geom_bar(color = 'black', fill = '#F79420') 

```

The data in the quality variable is normally distributed with following 
statistics.

```{r, echo = FALSE}
var = win_data$quality
summary(var)
```
Statistics of quality variable show that the majority of white wines fall into 
5 and 6 quality categories.


### Bivariate Plots Section and Analysis

From the uni-variant analysis in the above section, reflects some insights of 
individual attributes in the dataset. In this section, correlations between the 
attributes were investigated. 

In order to understand the relationships between the attributes, following 
correlation plots matrix was generated. The matrix of plots are advantageous
for these type of analysis,because, all the relationships between variables can
be easily identified. It is helpful to implement further analysis of the 
dataset. 

```{r, echo = FALSE, fig.height = 9, fig.width = 9, message = FALSE}
## Creating a correlation matrix plot

ggpairs(win_data)

# ggpairs(win_data[2:13], upper = list(continuous = "cor", combo = "dot", 
#   discrete ="facetbar", na = "na"), lower = list(continuous = "points", 
#   combo = "facethist", discrete = "facetbar", na = "na"), 
#   diag = list(continuous =  "densityDiag", discrete = "barDiag", 
#        na = "naDiag"),
#   xlab = NULL, ylab = NULL, axisLabels = c("show", "internal", "none"))

```

In the above figure, the scatter plots each attributes with the others and the 
corresponding correlation coefficients are shown.Along the diagonal of the 
diagram the density plots are displayed for each attribute. 


```{r, echo = FALSE, fig.height = 9, fig.width = 9, message = FALSE}
#coreletion = cor(win_data[2:12])
#coreletion = coreletion[coreletion > 0.5]
#coreletion
```
When examine the correlation coefficient for each relationship in above figure, 
some of the attributes seems to be high to moderately correlated to each other. 
For example, correlation coefficients of, alcohol vs density, residual sugar 
vs density, total sulfur dioxide vs density, total sulfur dioxide vs free sulfur 
dioxide are -0.78, 0.84, 0.53 and 0.62, respectively.

In the below sections, each of those high to moderate relationships were 
investigated after removing outliers if any. 

After preliminary removing density outliers, in alcohol % vs density and density 
vs total sulfur dioxide relationships, the correlation coefficients are slightly 
improved (-0.8 and 0.54, respectively). For the other relationships, taking 
density outliers off, did not improve the correlation (refer multivariate plots 
below to see the newly calculated correlation coefficients).

Following box plots were generated to further understand the behaviors of 
high to moderate correlated variables with the wine quality.


```{r, echo = FALSE, fig.height = 9, fig.width = 9, message = FALSE}
## Function to create a histogram
create_boxplot = function(dat_set, x = 'quality', y, x_lab, y_lab, 
                          unit = 'g/dm^3') {
    
    
                    return( ggplot(data = dat_set, aes_string(x = x, y = y) ) + 
                            geom_boxplot() +
                            xlab(x_lab) +
                            ylab(paste(y_lab, unit)) +
                            ggtitle(paste(x_lab, 'vs', y_lab)) +
                            theme(plot.title = element_text(size=11, hjust=0.5,
                                                            face = "bold"),
                            axis.title = element_text(size=10, hjust=0.5)))
                }
```

```{r, echo = FALSE, fig.height = 9, fig.width = 9, message = FALSE}
## Generating bi-variate plots after removing the outliers

win_data_w = win_data[win_data$total.sulfur.dioxide < 300,]

p11 = create_boxplot(win_data_z, 'quality', 'density', 'Quality', 'Density', 
                     '(g/cm^3)')

p12 = create_boxplot(win_data, 'quality', 'alcohol', 'Quality', 'Alcohol %', '')

p13 = create_boxplot(win_data_x, 'quality', 'residual.sugar', 'Quality', 
               'Residual sugar concentration')
p14 = create_boxplot(win_data_w, 'quality', 'total.sulfur.dioxide', 'Quality', 
               'Total sulfur dioxide')

p15 = create_boxplot(win_data_y, 'quality', 'free.sulfur.dioxide', 'Quality', 
               'Free sulfur dioxide')


grid.arrange(p11, p12, p13, p14, p15, ncol = 2)


```

After removing outliers identified in univariate section, data in these box 
plots indicate still there are more outliers can be present in selected 
variables (the black color points shows potential outliers). To remove these
outliers further understanding the dataset needed.



### Multivariate Plots Section and Analysis

Now its time to check how these relationships (shown in the above bi-variant 
plots) affect to the wine quality. In the following plots, the attributes are 
differentiating by the 'quality' variable (different color for each quality 
category).

The quality variable was grouped into 3 main categories as follows in order to
see the differences clearly (created a new variable 'Rank'). 

* Quality 1 to 4 = low quality
* Quality 5 to 6 = medium quality
* Quality 7 to 10 = high quality

```{r, echo = FALSE, fig.height = 9, fig.width = 9, message = FALSE}
## Creating a new variable called Rank.
win_data$Rank =  ifelse(win_data$quality == '3', 'low',
                        ifelse(win_data$quality == '4', 'low',
                        ifelse(win_data$quality == '5', 'medium',
                        ifelse(win_data$quality == '6', 'medium',
                        ifelse(win_data$quality == '7', 'high',
                        ifelse(win_data$quality == '8', 'high',
                        ifelse(win_data$quality == '9', 'high',
                        ifelse(win_data$quality == '10','high','low')))))))) 

## Recalculate with the new variable
## Remove the density > 1.00
win_data1 = win_data[win_data$density < 1.00,]
## Remove extreme values of total sulfur dioxide
win_data2 = win_data1[win_data1$total.sulfur.dioxide < 300,]
## Remove extreme values of free sulfur dioxide
win_data3 = win_data2[win_data2$free.sulfur.dioxide < 300,]
```



```{r, echo = FALSE, fig.height = 9, fig.width = 9, message = FALSE}

## Generating multi-variate plots grouping by rank
x = win_data1$alcohol
y = win_data1$density
p1 = ggplot(aes(x = alcohol, y = density),  data = win_data1) + 
        geom_point(aes(colour = Rank), alpha = 1/3, 
                   position = position_jitter(h = 0))+
        ylab("Density (g/cm^3)") +
        xlab("Alcohol %") +
        ggtitle("Alcohol % vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

x = win_data1$residual.sugar
y = win_data1$density
p2 = ggplot(aes(x = residual.sugar, y = density ), data = win_data1) + 
        geom_point(aes(colour = Rank), alpha = 1/3, 
                   position = position_jitter(h = 0))+
        ylab("Density (g/cm^3)") +
        xlab("Residual sugar concentration (g/dm^3)") +
        ggtitle(" Residual sugar vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))


x = win_data2$total.sulfur.dioxide
y = win_data2$density

p3 = ggplot(aes(x = total.sulfur.dioxide, y = density), data = win_data2) + 
        geom_point(aes(colour = Rank),alpha = 1/3, 
                   position = position_jitter(h = 0)) +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Density (g/cm^3)") +
        ggtitle("Total sulfur dioxide vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

y = win_data3$free.sulfur.dioxide 
x = win_data3$total.sulfur.dioxide
p4 = ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide), 
                data = win_data3) + 
        geom_point(aes(colour = Rank), alpha = 1/3, 
                position = position_jitter(h = 0)) +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Free sulfur dioxide (g/dm^3)") +
        ggtitle("Total vs free sulfur dioxide", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+ 
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

grid.arrange(p1, p2, p3, p4, ncol = 2)


```


The above four plots show, how the quality (Rank) of wine behave with different
attributes. The observations of the plots are discussed below except for the 
plot total vs free sulfur dioxide concentrations, due to absence of identifiable 
patterns in that particular plot.


### Final Plots and Summary

#### Behaviours of the quality of wines related to the changes of alcohol % and density


```{r, echo = FALSE, Plot_One}
## reprint the plot
p1
```

Based on the data in the above plot, the data points corresponding to the 
high quality (red color points) mainly clustered towards the bottom right side 
of the plot. This behavior indicates the quality of wine is generally high when
the alcohol percentage is high and density is low with a few exceptions. 





#### Behaviours of the quality of wines related to the changes of residual sugar concentration and density 
```{r, echo = FALSE, Plot_Two}
## reprint the plot
p2
```

In above plot, there is slight separation of high quality wines with the other 
qualities. That arrangement of high quality data points suggest that in general 
the quality of a wine type is high when the density and the residual sugar 
concentrations are low with some exceptions.





#### Behaviours of the quality of wines related to the changes of total sulfur dioxide concentration and density

```{r, echo = FALSE, Plot_Three}
## reprint the plot
p3 

```

Similar to previous two plots, this plot also moderately differentiate high 
quality wines with the others. Generally, the low density and somewhat sulfur 
dioxide content (about 75 to 125 g/dm^3) give high quality wines.


#### Creating a Multiple linear regression model

During above multi-variant analysis, it was noted that density plays major role 
of determining wine quality with variation of the other parameters such as sugar
content, alcohol percentage and sulfur dioxide content. Therefore, multiple 
regression analysis was carried out to determine a relationship, if any, 
between density of wines and other three parameters. 

The following is the summary of the multiple linear regression model generated. 
The summary include the correlation coefficient (0.92), y-intercept (1.00), and 
slopes for each parameter for the model along with other important parameters. 


```{r, echo = FALSE, multiple_linearr_regression0}
# Multiple Linear Regression model
fit <- lm(density ~ total.sulfur.dioxide + residual.sugar + 
              alcohol, data=win_data3)
summary(fit) # show the summary of results

#coefficients(fit) # model coefficients
#confint(fit, level=0.95) # CIs for model parameters 
#fitted(fit) # predicted values
#residuals(fit) # residuals
#anova(fit) # anova table 
#vcov(fit) # covariance matrix for model parameters 
#influence(fit) # regression diagnostics
```

In this multiple linear regression model, the correlation coefficient of the 
density vs other selected parameters improved to 0.92 from the previous maximum 
individual parameter coefficient of 0.82. 

All the p-values for model parameters (y-intercept and slopes) are less than 
0.05. Therefore, the model parameters are statistically significant. That 
indicates the robustness of the model.

 
#### Validating the model

We can determine the validity of the model by analyzing the following 
characterization plots and the histograms in the uni-variant section.

The main assumptions for validating a linear model are as follows.

* The variables should be independent from each other.
* The individual variables should be normally distributed.
* The residuals should not show specific patterns.

Since, all of these assumptions are fulfilled by the selected variables in the 
model, the validity of the model is confirmed.


```{r, echo = FALSE, multiple_linearr_regression1}
## getting parameters to validate the model
par(mfrow =c(2,2))
plot(fit)
```

#### Displaying the model with calculated and measured wine density data

The following plot shows the predicted densities using the regression model 
derived in above section.

```{r, echo = FALSE, multiple_linearr_regression2}
## generating a new variable (predicted density)
intercept =  1.004e+00
coef1 =  5.861e-06  
coef2 =  3.233e-04  
coef3 = -1.192e-03


win_data3$predict_density = win_data3$total.sulfur.dioxide*(coef1) +
    win_data3$residual.sugar*(coef2) +
    win_data3$alcohol*(coef3) +intercept


```

```{r, echo = FALSE, multiple_linearr_regression3}
## Generating simple linear regression model
fit2 = lm(predict_density ~ density, data = win_data3)

## Plot the model
ggplot(aes(density , predict_density), data = win_data3) +
    geom_point(aes(color= Rank), alpha = 1/3, position =
                   position_jitter(h = 0)) +
    geom_abline(intercept = coef(fit2)[1], slope = coef(fit2)[2], col='blue')+
    xlab("Density (g/cm^3)") +
    ylab("Predicted density (g/cm^3)") 

#coef(fit2)

```

In this plot, the multiple linear regression model is shown in dark blue line.
The data points are the measured densities and the calculated densities for the 
wines.In addition, the data points are marked by the quality of wines as 
different colors.Based on this model, the wine quality can be determined for 
most of the wines by their density. However, more detailed analysis and 
classification such as machine learning techniques should be carried out to 
improve the model and to increase the accuracy of the model. 


### Reflection

A data set related to white wine quality was investigated and analysed. By 
definition, some of the variables have similar characteristics and relation.
For example, pH and citric acid concentration should be related each other, 
because, literally those variable reflects same information. However, when 
determining the relationships of the variables each other (see the bi-variant 
matrix of plots) pH and citric acid concentration do not show a good 
relationship as expected (R^2 = -0.16). 

Density of seems to have better correlation between sugar, alcohol and sulfur 
dioxide content. Based on the these relationships, using density and other 
mentioned variables, a multiple linear regression model was generated. This 
generated model can be used to determine the wine quality. However, the model 
needs to be improved to get accurate predictions. Other classification 
techniques such that techniques in The machine-leaning methods can be used to 
improve this model or define a separate more accurate model. Some discrepancies
of data quality observed, therefore, the validity and quality of the data 
should be examined before doing further analysis with the dataset.


