Lasantha Rathnayake
========================================================
## Analysis of white wine quality
In this exploratory data analysis study, white wine Quality is discussed.The 
dataset used in this investigation is 'White Wine Quality', which is a part of 
two datasets for white wine and red whine. Following is the reference for the 
dataset.

*   P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
    Modeling wine preferences by data mining from physicochemical properties.
    In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
## Installing required packages
# install.packages("ggplot2", dependencies = T) 
# install.packages("knitr", dependencies = T)
# install.packages("dplyr", dependencies = T)
#install.packages('GGally', dependencies = T)

setwd("E:/OneDrive_gmail/OneDrive/Dropbox_transfers/Udacity/Nano_Degree_Data_analysis/Project_6_R_programing/R_final_project")
## Load required packages
library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(GGally)

```


### Information about dataset

In the dataset, there are over 4800 records are available in 11 attributes. Most 
of the attributes are the chemical and physical properties of the wine type.

The dataset contains following attributes.
```{r echo=FALSE, Load_the_Data}
# Load the Data
#getwd()
#list.files()
win_data =  read.csv('wineQualityWhites.csv', sep = ',')
names(win_data)

```
The size of the dataset is as follows.
```{r, echo=FALSE}
dim(win_data)
```
Based on the reference of the dataset, the following descriptions are given for 
the attributes in the dataset. The corresponding units if any for the atributes 
are also given in along with the atribute name.

fixed acidity (g/dm^3):

*   Most acids involved with wine or fixed or nonvolatile 
    (do not evaporate readily).

volatile acidity (g/dm^3):

*   The amount of acetic acid in wine, which at too high of levels can lead to 
    an unpleasant, vinegar taste

citric acid (g/dm^3):

*   Found in small quantities, citric acid can add 'freshness' and flavor 
    to wines

residual sugar (g/dm^3):

*   The amount of sugar remaining after fermentation stops, it's rare to 
    find wines with less than 1 gram/liter and wines with greater than 
    45 grams/liter are considered sweet

chlorides (g/dm^3): 

*   The amount of salt in the wine

free sulfur dioxide (g/dm^3):

*   The free form of SO2 exists in equilibrium between molecular SO2 
    (as a dissolved gas) and bisulfite ion; it prevents microbial 
    growth and the oxidation of wine

total sulfur dioxide (g/dm^3):

*   Amount of free and bound forms of S02; in low concentrations, SO2 is mostly 
    undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 
    becomes evident in the nose and taste of wine

density (g/cm^3): 

*   The density of water is close to that of water depending on the percent 
    alcohol and sugar content

pH: 
*   Describes how acidic or basic a wine is on a scale from 0 (very acidic) to 
    14 (very basic); most wines are between 3-4 on the pH scale

sulphates (g/dm^3):

*   A wine additive which can contribute to sulfur dioxide gas (S02) levels, 
    wich acts as an antimicrobial and antioxidant

alcohol (% by volume): 

*   The percent alcohol content of the wine

quality: 

*   Score between 0 and 10



The following snippet of data shows the first six records of the dataset.

```{r, echo=FALSE}
head(win_data)
```

A summary of the main attributes are given below. 

```{r, echo=FALSE}
#summary(win_data[2:12])

```





### Univariate Plots Section

```{r, echo=FALSE, fig.height= 7}

## Making histograms and barcharts for the main atributes
p1 = ggplot(aes(x = fixed.acidity), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .25, color = 'black', 
                       fill = '#F79420')

p2 = ggplot(aes(x = volatile.acidity), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420') 

p3 = ggplot(aes(x = citric.acid), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420') 

p4 = ggplot(aes(x = residual.sugar), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 1, color = 'black', 
                       fill = '#F79420')

p5 = ggplot(aes(x = chlorides), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .007, color = 'black', 
                       fill = '#F79420') 

p6 = ggplot(aes(x = free.sulfur.dioxide), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 6, color = 'black', 
                       fill = '#F79420')

p7 =  ggplot(aes(x = total.sulfur.dioxide), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 8, color = 'black', 
                       fill = '#F79420')

p8 = ggplot(aes(x = density), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .001, color = 'black', 
                       fill = '#F79420')

p9 =  ggplot(aes(x = pH), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420')

p10 = ggplot(aes(x = sulphates), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .02, color = 'black', 
                       fill = '#F79420')

p11 = ggplot(aes(x = alcohol), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .25, color = 'black', 
                       fill = '#F79420') 

p12 = ggplot(aes(x = quality), data = win_data) + 
        geom_bar(color = 'black', fill = '#F79420') 
    

grid.arrange(p1, p2, p3,p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol = 3)

```


All the atributes in the dataset are numerical. However, The quality attribute 
was factorized, because the quality atribute is used to rank the qualty.


### Univariate Analysis

Based on the above grid of histograms and barchart, it is obvious that all the 
measured parameters for white wine are more or less normally distributed except 
'residual sugar' which is left skewed. The distribution of pH is perfect normal distribution. 

The following summary shows the main descriptive statistics of the atributes.


```{r, echo=FALSE}
win_data[, 'quality'] <- as.factor(win_data[, 'quality'])
summary(win_data[2:13])

```


Based on the above summary statistics, some outlier like data points can be observed specially for the atributes - residual.sugar, free.sulfur.dioxide and total.sulfur.dioxide. In this three atributes, the maximum value are 
significantly higher than the corresponding third quartile. However, to veryfy whether these max pointes are outlieers or not, a detailed analysis are required. Also, the stitistcs show that the majority of white wines fall into 5 and 6 quality categories.  



### Bivariate Plots Section

From the univariant analysis in the above section reflects some insights of individual atributes in the dataset. In this section correlations between the atributes are investigated. 

In order to understand the relationships between the atributes, following coorellation plots matrixs was generated. The matrix of plots are advantageous, because, all the relationships if any can be eaily identified. Therefore, further analysis can be easily implemented. 

```{r echo=FALSE, fig.height=9, fig.width=9, warning=FALSE, message=FALSE}

ggpairs(win_data[2:13]) 

#ggpairs(win_data[2:13], upper = list(continuous = "cor", combo = "dot", discrete ="facetbar", na = "na"), lower = list(continuous = "points", combo = "facethist", discrete = "facetbar", na = "na"), diag = list(continuous =  "densityDiag", discrete = "barDiag", na = "naDiag"), xlab = NULL, ylab = NULL, axisLabels = c("show", "internal", "none"))

```

In the above matrix of plots, scatter plots were shown each atributes to others 
at the lower part of the diogonal. I the upper part, the corresponding 
correlation coefficients are shown. Along the diogonal the density plots are 
displayed for each atribute. 



### Bivariate Analysis
```{r,echo=FALSE, fig.height=9, fig.width=9, warning=FALSE, message=FALSE}
#coreletion = cor(win_data[2:12])
#coreletion = coreletion[coreletion > 0.5]
#coreletion
```
When examine the correlation coeffient for each relationship, some of the 
atributes seems to be high to moderately coorelated each other. For example, 
corellation coefficients of, alcohol vs density, residual.sugar vs density, total
.sulfur.dioxide vs density, total.sulfur.dioxide vs free.sulfur.dioxide are 
-0.78, 0.84, 0.53 and 0.62, repectively.

Let's look at each relationship seperately.

```{r,echo=FALSE,  warning=FALSE, message=FALSE, fig.height=5}

x = win_data$alcohol
y = win_data$density
p1 = ggplot(aes(x = alcohol, y = density ), data = win_data) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Alcohol %") +
        ggtitle("Alcohol % vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

x = win_data$residual.sugar
y = win_data$density
p2 = ggplot(aes(x = residual.sugar, y = density ), data = win_data) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Residual sugar concentration (g/dm^3)") +
        ggtitle(" Residual sugar vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))
x = win_data$total.sulfur.dioxide
y = win_data$density

p3 = ggplot(aes(x = total.sulfur.dioxide, y = density), data = win_data) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Density (g/cm^3)") +
        ggtitle("Total sulfur dioxide vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))


y = win_data$free.sulfur.dioxide 
x = win_data$total.sulfur.dioxide
p4 = ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide), data = win_data) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Free sulfur dioxide (g/dm^3)") +
        ggtitle("Total vs free sulfur dioxide", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+ 
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

grid.arrange(p1, p2, p3, p4, ncol = 2)


```

The data distribution in above plots shows faily good correlations with R^2 > 0.5.

As noticed in the univarient analysis in the previous section, there seems to be
a few outliers observable in the datset. Generally, in standaed conditions the 
density of water is 1.00 g/cm^3. Therefore, the density of a solution in which 
water is the solvent, should be slightly less than 1.00 g/cm^3. In the above 
plots we can observe two density values greater tahn 1.00 g/cm^3. Therefore, we 
can remove those outliers from the dataset. Folowing plots were generated without
the density outliers. Also, extreame valus for sulfur dioxides can be removed.


```{r,echo=FALSE,  warning=FALSE, message=FALSE, fig.height=5}

win_data1 = win_data[win_data$density < 1.00,]

x = win_data1$alcohol
y = win_data1$density
p1 = ggplot(aes(x = alcohol, y = density ), data = win_data1) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Alcohol %") +
        ggtitle("Alcohol % vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

x = win_data1$residual.sugar
y = win_data1$density
p2 = ggplot(aes(x = residual.sugar, y = density ), data = win_data1) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Residual sugar concentration (g/dm^3)") +
        ggtitle(" Residual sugar vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

win_data2 = win_data1[win_data1$total.sulfur.dioxide < 300,]
x = win_data2$total.sulfur.dioxide
y = win_data2$density

p3 = ggplot(aes(x = total.sulfur.dioxide, y = density), data = win_data2) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Density (g/cm^3)") +
        ggtitle("Total sulfur dioxide vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

win_data3 = win_data2[win_data2$free.sulfur.dioxide < 200,]

y = win_data3$free.sulfur.dioxide 
x = win_data3$total.sulfur.dioxide
p4 = ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide), data = win_data3) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Free sulfur dioxide (g/dm^3)") +
        ggtitle("Total vs free sulfur dioxide", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+ 
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

grid.arrange(p1, p2, p3, p4, ncol = 2)


```

Now the data distrbution and the visuabilty of the plots are improved. In adition,
in alcohol % vs density and density vas total sulfur dioxide relationships, 
correlation coeeficient are sligtly improved. For the other relationships, 
taking outliers off, did not improve the correlation.



### Multivariate Plots Section

Now its time to check how these relationships (in above bivarient plots) affect
to the wine quality. Folowwing plots difrentshiate the atributes by quality 
(diffrent colr for each quality category).

In this case the quality is group into 3 main categories as follows in order to
see the differences clearly. 

* Quality 1 to 4 = low quality
* Quality 5 to 6 = medium quality
* Quality 7 to 10 = high qualiy

```{r,echo=FALSE,  warning=FALSE, message=FALSE, fig.height=6}
win_data$Rank =  ifelse(win_data$quality == '3', 'low',
                        ifelse(win_data$quality == '4', 'low',
                        ifelse(win_data$quality == '5', 'medium',
                        ifelse(win_data$quality == '6', 'medium',
                        ifelse(win_data$quality == '7', 'high',
                        ifelse(win_data$quality == '8', 'high',
                        ifelse(win_data$quality == '9', 'high',
                        ifelse(win_data$quality == '10','high','low')))))))) 

```

```{r,echo=FALSE,  warning=FALSE, message=FALSE, fig.height=6}

win_data1 = win_data[win_data$density < 1.00,]

x = win_data1$alcohol
y = win_data1$density
p1 = ggplot(aes(x = alcohol, y = density),  data = win_data1) + 
        geom_point(aes(colour = Rank), alpha = 1/3, position = position_jitter(h = 0))+
        ylab("Density (g/cm^3)") +
        xlab("Alcohol %") +
        ggtitle("Alcohol % vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

x = win_data1$residual.sugar
y = win_data1$density
p2 = ggplot(aes(x = residual.sugar, y = density ), data = win_data1) + 
        geom_point(aes(colour = Rank), alpha = 1/3, position = position_jitter(h = 0))+
        ylab("Density (g/cm^3)") +
        xlab("Residual sugar concentration (g/dm^3)") +
        ggtitle(" Residual sugar vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

win_data2 = win_data1[win_data1$total.sulfur.dioxide < 300,]
x = win_data2$total.sulfur.dioxide
y = win_data2$density

p3 = ggplot(aes(x = total.sulfur.dioxide, y = density), data = win_data2) + 
        geom_point(aes(colour = Rank),alpha = 1/3, position = position_jitter(h = 0)) +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Density (g/cm^3)") +
        ggtitle("Total sulfur dioxide vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

win_data3 = win_data2[win_data2$free.sulfur.dioxide < 200,]

y = win_data3$free.sulfur.dioxide 
x = win_data3$total.sulfur.dioxide
p4 = ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide), data = win_data3) + 
        geom_point(aes(colour = Rank), alpha = 1/3, position = position_jitter(h = 0)) +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Free sulfur dioxide (g/dm^3)") +
        ggtitle("Total vs free sulfur dioxide", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+ 
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

grid.arrange(p1, p2, p3, p4, ncol = 2)


```

The above four plots show, how the quality (Rank) of wine behave with diffrent
parameters. The observations of the plots are given below except the bottom right
plot (Total vs free sulfur dioxide concentrations) where, there is no udentifiable 
patterns in the wine quality with corresponding to x and y axis parameters 
changing. In the other plots there are visible wine quality patters can be seen. 



### Multivariate Analysis


#### Behaviours of the quality of wines related to the changes of alcohol % and density
```{r echo=FALSE, Plot_One}
p1
```

Based on the data in the top above plot, the data poins corresponding to the 
high quality mainly clusterd towards botom right side of the plot. This behaviour
indicates the quality of wine is generaly high when the alcohol percentage is 
high and denity is low. 



#### Behaviours of the quality of wines related to the changes of residual suger concentration and density 
```{r echo=FALSE, Plot_Two}
p2
```

In above plot, there is slight seperation of high quality wines with the other 
qualities. That arrangement of hiquality data poits suggest that in genaral the 
quality of a wine is high when the density and the residual sugar concentrationa 
are low.



#### Behaviours of the quality of wines related to the changes of total sulfur dioxide concentration and density

```{r echo=FALSE, Plot_Three}
p3 

```

Similar to previuos two plots, this plot also moderately diifrentiate high quality
wines with the others. Generally, the low density and somewhat sulfur dioxide 
(about 75 to 125 g/dm^3) gives high quality wines.


### Creating a Multiple linear regression model

During above mutivariant analysis, it was noted that density plays major role of
detrming wine quality with variation other parameters such as suger content, 
alcohol percentage and sulfur dioxide content. Therefore, multiple regression 
analysis carried out to detrmine the relationship between density of wines and
other three parameters. 

The follwing is the summary of the multiple linear regression medel. The summry 
include the corelation coeeficient (0.92), y-intercept (1.00), and slopes for each parameter
are given for the model along with the other important parametrs. 


```{r echo=FALSE, multiple_linearr_regression0}
# Multiple Linear Regression 
fit <- lm(density ~ total.sulfur.dioxide + residual.sugar + alcohol, data=win_data3)
summary(fit) # show the summary of results

#coefficients(fit) # model coefficients
#confint(fit, level=0.95) # CIs for model parameters 
#fitted(fit) # predicted values
#residuals(fit) # residuals
#anova(fit) # anova table 
#vcov(fit) # covariance matrix for model parameters 
#influence(fit) # regression diagnostics
```

In this multiple linear regression model, the correlation of the deisity vs other
selected parameters improved to 0.92 from prvious maximum coeficent 0.82. 
Sice the model is improved with the selected parametrs, one can suggest that
density can be explaned alcohol, suger and sulfur dioxide content. All the 
p-values for model parametrs (y-intercep and slopes for each variables) are less
than 0.05, the model is statisticaly significant.

Further we can determine the validity of the model by fillowing charaterization 
plots and the histograms from the univarient sections. The main assumptions such 
that:

* The variables indipendant each other
* The indivdual variables are normaly distributed
* The residuals do not show specific pattrens. etc.


```{r echo=FALSE, multiple_linearr_regression1}
par(mfrow =c(2,2))
plot(fit)
```

The following plot shows the predicted densities using the regression model 
derived in above section.

```{r echo=FALSE, multiple_linearr_regression2}
## generating a new variable (predicted density)
intercept =  1.004e+00
coef1 =  5.861e-06  
coef2 =  3.233e-04  
coef3 = -1.192e-03


win_data3$predict_density = win_data3$total.sulfur.dioxide*(coef1) +
    win_data3$residual.sugar*(coef2) +
    win_data3$alcohol*(coef3) +intercept


```

```{r echo=FALSE, multiple_linearr_regression3}
fit2 = lm(predict_density ~ density, data = win_data3)
ggplot(aes(density , predict_density), data = win_data3) +
    geom_point(aes(color= Rank), alpha = 1/3, position =
                   position_jitter(h = 0)) +
    geom_abline(intercept = coef(fit2)[1], slope = coef(fit2)[2], col='blue') +
    xlab("Density (g/cm^3)") +
    ylab("Predicted density (g/cm^3)") 

#coef(fit2)

```

In above plot, the multiple linear regression model is shown in dark blue line. 
In addition the data points are marked by the quality of wines as diffrent colors.
Based on this model, the wine quality can be determined for most of the wines 
by their density. However, more detailed analysis and classification such as 
macine learning techniques needed to improve this model and to increse the 
accuracy, because, still a considerable amount of data points related to high
and low quality mixed together.


### Reflection

A data set related to white wine quality was investigated and analysed. By 
definitio some of the variable has similar charateristics. For example, pH and 
citric acid concentration. However, when determoning the relationships of the
variables each other (see the bivarient matrix of plots) pH and ctric adid 
concentration do not show a good relationship as expected (R^2 = -0.16). 
Density of semms to have better corelatons between suger, alcohol and sulfur 
dioxide content. Based on the these relationships with density and other mentioned 
variables, a multiple linear regression model was generated. This generated 
model can be used to determoine the wine quality. However, the model needs to be
improved to get accurate predictions. Other classification techniques such that 
in The machine leaning techniques can be used to improve this model or define
a more accurate model. Some decrepancies of data quality observed, therefore, 
the validity and quality of the data should be examined before doing further 
analysis with the dataset.


