Lasantha Rathnayake
========================================================
## Analysis of white wine quality
In this exploratory data analysis study, white wine Quality is discussed.The 
dataset used in this investigation is 'White Wine Quality', which is a part of 
two datasets for white wine and red whine. Following is the reference for the 
dataset.

*   P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
    Modeling wine preferences by data mining from physicochemical properties.
    In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
## Installing required packages
# install.packages("ggplot2", dependencies = T) 
# install.packages("knitr", dependencies = T)
# install.packages("dplyr", dependencies = T)
#install.packages('GGally', dependencies = T)

setwd("E:/OneDrive_gmail/OneDrive/Dropbox_transfers/Udacity/Nano_Degree_Data_analysis/Project_6_R_programing/R_final_project")
## Load required packages
library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(GGally)

```
### Information about dataset

In the dataset, there are over 4800 records are available in 11 attributes. Most 
of the attributes are the chemical and physical properties of the wine type.

The dataset contains following attributes.
```{r echo=FALSE, Load_the_Data}
# Load the Data
#getwd()
#list.files()
win_data =  read.csv('wineQualityWhites.csv', sep = ',')
names(win_data)

```
The size of the dataset is as follows.
```{r, echo=FALSE}
dim(win_data)
```
Based on the reference of the dataset, the following descriptions are given for 
the attributes in the dataset. The corresponding units if any for the atributes 
are also given in along with the atribute name.

fixed acidity (g/dm^3):

*   Most acids involved with wine or fixed or nonvolatile 
    (do not evaporate readily).

volatile acidity (g/dm^3):

*   The amount of acetic acid in wine, which at too high of levels can lead to 
    an unpleasant, vinegar taste

citric acid (g/dm^3):

*   Found in small quantities, citric acid can add 'freshness' and flavor 
    to wines

residual sugar (g/dm^3):

*   The amount of sugar remaining after fermentation stops, it's rare to 
    find wines with less than 1 gram/liter and wines with greater than 
    45 grams/liter are considered sweet

chlorides (g/dm^3): 

*   The amount of salt in the wine

free sulfur dioxide (g/dm^3):

*   The free form of SO2 exists in equilibrium between molecular SO2 
    (as a dissolved gas) and bisulfite ion; it prevents microbial 
    growth and the oxidation of wine

total sulfur dioxide (g/dm^3):

*   Amount of free and bound forms of S02; in low concentrations, SO2 is mostly 
    undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 
    becomes evident in the nose and taste of wine

density (g/cm^3): 

*   The density of water is close to that of water depending on the percent 
    alcohol and sugar content

pH: 
*   Describes how acidic or basic a wine is on a scale from 0 (very acidic) to 
    14 (very basic); most wines are between 3-4 on the pH scale

sulphates (g/dm^3):

*   A wine additive which can contribute to sulfur dioxide gas (S02) levels, 
    wich acts as an antimicrobial and antioxidant

alcohol (% by volume): 

*   The percent alcohol content of the wine

quality: 

*   Score between 0 and 10



The following snippet of data shows the first six records of the dataset.

```{r, echo=FALSE}
head(win_data)
```

A summary of the main attributes are given below. 

```{r, echo=FALSE}
#summary(win_data[2:12])

```





### Univariate Plots Section

```{r, echo=FALSE, fig.height= 7}

## Making histograms and barcharts for the main atributes
p1 = ggplot(aes(x = fixed.acidity), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .25, color = 'black', 
                       fill = '#F79420')

p2 = ggplot(aes(x = volatile.acidity), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420') 

p3 = ggplot(aes(x = citric.acid), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420') 

p4 = ggplot(aes(x = residual.sugar), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 1, color = 'black', 
                       fill = '#F79420')

p5 = ggplot(aes(x = chlorides), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .007, color = 'black', 
                       fill = '#F79420') 

p6 = ggplot(aes(x = free.sulfur.dioxide), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 6, color = 'black', 
                       fill = '#F79420')

p7 =  ggplot(aes(x = total.sulfur.dioxide), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 8, color = 'black', 
                       fill = '#F79420')

p8 = ggplot(aes(x = density), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .001, color = 'black', 
                       fill = '#F79420')

p9 =  ggplot(aes(x = pH), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420')

p10 = ggplot(aes(x = sulphates), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .02, color = 'black', 
                       fill = '#F79420')

p11 = ggplot(aes(x = alcohol), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .25, color = 'black', 
                       fill = '#F79420') 

p12 = ggplot(aes(x = quality), data = win_data) + 
        geom_bar(color = 'black', fill = '#F79420') 
    

grid.arrange(p1, p2, p3,p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol = 3)

```


All the atributes in the dataset are numerical. However, The quality attribute 
was factorized, because the quality atribute is used to rank the qualty.

### Univariate Analysis

Based on the above grid of histograms and barchart, it is obvious that all the 
measured parameters for white wine are more or less normally distributed except 
'residual sugar' which is left skewed. The distribution of pH is perfect normal distribution. 

The following summary shows the main descriptive statistics of the atributes.


```{r, echo=FALSE}
win_data[, 'quality'] <- as.factor(win_data[, 'quality'])
summary(win_data[2:13])

```


Based on the above summary statistics, some outlier like data points can be observed specially for the atributes - residual.sugar, free.sulfur.dioxide and total.sulfur.dioxide. In this three atributes, the maximum value are 
significantly higher than the corresponding third quartile. However, to veryfy whether these max pointes are outlieers or not, a detailed analysis are required. Also, the stitistcs show that the majority of white wines fall into 5 and 6 quality categories.  


### Bivariate Plots Section

From the univariant analysis in the above section reflects some insights of individual atributes in the dataset. In this section correlations between the atributes are investigated. 

In order to understand the relationships between the atributes, following coorellation plots matrixs was generated. The matrix of plots are advantageous, because, all the relationships if any can be eaily identified. Therefore, further analysis can be easily implemented. 

```{r echo=FALSE, fig.height=9, fig.width=9, warning=FALSE, message=FALSE}

ggpairs(win_data[2:13]) 

#ggpairs(win_data[2:13], upper = list(continuous = "cor", combo = "dot", discrete ="facetbar", na = "na"), lower = list(continuous = "points", combo = "facethist", discrete = "facetbar", na = "na"), diag = list(continuous =  "densityDiag", discrete = "barDiag", na = "naDiag"), xlab = NULL, ylab = NULL, axisLabels = c("show", "internal", "none"))

```

In the above matrix of plots, scatter plots were shown each atributes to others 
at the lower part of the diogonal. I the upper part, the corresponding 
correlation coefficients are shown. Along the diogonal the density plots are 
displayed for each atribute.    

### Bivariate Analysis
```{r,echo=FALSE, fig.height=9, fig.width=9, warning=FALSE, message=FALSE}
#coreletion = cor(win_data[2:12])
#coreletion = coreletion[coreletion > 0.5]
#coreletion
```
When examine the correlation coeffient for each relationship, some of the 
atributes seems to be high to moderately coorelated each other. For example, 
corellation coefficients of, alcohol vs density, residual.sugar vs density, total
.sulfur.dioxide vs density, total.sulfur.dioxide vs free.sulfur.dioxide are 
-0.78, 0.84, 0.53 and 0.62, repectively.

Let's look at each relationship seperately.

```{r,echo=FALSE,  warning=FALSE, message=FALSE, fig.height=5}

p1 = ggplot(aes(x = alcohol, y = density ), data = win_data) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Alcohol %") +
        ggtitle("Alcohol % vs density: R^2= -0.78") + 
        theme(plot.title = element_text(size=10, hjust=0.5),
              axis.title = element_text(size=10, hjust=0.5))


p2 = ggplot(aes(x = residual.sugar, y = density ), data = win_data) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Residual sugar concentration (g/dm^3)") +
        ggtitle(" Residual sugar vs Wine: R^2= 0.84")+ 
        theme(plot.title = element_text(size=10, hjust=0.5),
              axis.title = element_text(size=10, hjust=0.5))


p3 = ggplot(aes(x = total.sulfur.dioxide, y = density), data = win_data) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Density (g/cm^3)") +
        ggtitle("Total sulfur dioxide vs density: R^2= 0.53")+ 
        theme(plot.title = element_text(size=10, hjust=0.5),
              axis.title = element_text(size=10, hjust=0.5))



p4 = ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide), data = win_data) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Free sulfur dioxide (g/dm^3)") +
        ggtitle("Total vs free sulfur dioxide") + #, subtitle = "R^2= 0.62")+ 
        theme(plot.title = element_text(size=10, hjust=0.5),
              axis.title = element_text(size=10, hjust=0.5))

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The data distribution in above plots shows faily good correlations with R^2 > 0.5.

As noticed in the univarient analysis in the previous section, there seems to be
a few outliers observable in the datset. Generally, in standaed conditions the 
density of water is 1.00 g/cm^3. Therefore, the density of a solution in which 
water is the solvent, should be slightly less than 1.00 g/cm^3. In the above 
plots we can observe two density values greater tahn 1.00 g/cm^3. Therefore, we 
can remove those outliers from the dataset. Folowing plots were generated without
the density outliers. Also, extreame valus for sulfur dioxides can be removed.


```{r,echo=FALSE,  warning=FALSE, message=FALSE, fig.height=5}

win_data1 = win_data[win_data$density < 1.00,]

x = win_data1$alcohol
y = win_data1$density
p1 = ggplot(aes(x = alcohol, y = density ), data = win_data1) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Alcohol %") +
        ggtitle("Alcohol % vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

x = win_data1$residual.sugar
y = win_data1$density
p2 = ggplot(aes(x = residual.sugar, y = density ), data = win_data1) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Residual sugar concentration (g/dm^3)") +
        ggtitle(" Residual sugar vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))
x = win_data1$total.sulfur.dioxide
y = win_data1$density

p3 = ggplot(aes(x = total.sulfur.dioxide, y = density), data = win_data1) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Density (g/cm^3)") +
        ggtitle("Total sulfur dioxide vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

win_data2 = win_data[win_data$free.sulfur.dioxide < 200,]

y = win_data2$free.sulfur.dioxide 
x = win_data2$total.sulfur.dioxide
p4 = ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide), data = win_data2) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Free sulfur dioxide (g/dm^3)") +
        ggtitle("Total vs free sulfur dioxide", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+ 
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

grid.arrange(p1, p2, p3, p4, ncol = 2)


```



```{r,echo=FALSE,  warning=FALSE, message=FALSE, fig.height=5}

win_data1 = win_data[win_data$density < 1.00,]

x = win_data1$alcohol
y = win_data1$density
p1 = ggplot(aes(x = alcohol, y = density, color = quality ), data = win_data1) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Alcohol %") +
        ggtitle("Alcohol % vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

x = win_data1$residual.sugar
y = win_data1$density
p2 = ggplot(aes(x = residual.sugar, y = density, color = quality ), data = win_data1) + 
        geom_point()+
        ylab("Density (g/cm^3)") +
        xlab("Residual sugar concentration (g/dm^3)") +
        ggtitle(" Residual sugar vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))
x = win_data1$total.sulfur.dioxide
y = win_data1$density

p3 = ggplot(aes(x = total.sulfur.dioxide, y = density, color = quality), data = win_data1) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Density (g/cm^3)") +
        ggtitle("Total sulfur dioxide vs density", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+  
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

win_data2 = win_data[win_data$free.sulfur.dioxide < 200,]

y = win_data2$free.sulfur.dioxide 
x = win_data2$total.sulfur.dioxide
p4 = ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide, color = quality), data = win_data2) + 
        geom_point() +
        xlab("Total sulfur dioxide (g/dm^3)") +
        ylab("Free sulfur dioxide (g/dm^3)") +
        ggtitle("Total vs free sulfur dioxide", 
                subtitle = paste("R^2 = ", round(cor(x,y), digits = 2)))+ 
        theme(plot.title = element_text(size=11, hjust=0.5, face = "bold"),
              axis.title = element_text(size=10, hjust=0.5))

grid.arrange(p1, p2, p3, p4, ncol = 2)


```




> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!