Lasantha Rathnayake
========================================================
## Analysis of white wine quality
In this exploratory data analysis study, white wine Quality is discussed.The 
dataset used in this investigation is 'White Wine Quality', which is a part of 
two datasets for white wine and red whine. Following is the reference for the 
dataset.

*   P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
    Modeling wine preferences by data mining from physicochemical properties.
    In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
## Installing required packages
# install.packages("ggplot2", dependencies = T) 
# install.packages("knitr", dependencies = T)
# install.packages("dplyr", dependencies = T)

## Load required packages
library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)

```
### Information about dataset

In the dataset, there are over 4800 records are available in 11 attributes. Most 
of the attributes are the chemical and physical properties of the wine type.

The dataset contains following attributes.
```{r echo=FALSE, Load_the_Data}
# Load the Data
#getwd()
#list.files()
win_data =  read.csv('wineQualityWhites.csv', sep = ',')
names(win_data)

```
The size of the dataset is as follows.
```{r, echo=FALSE}
dim(win_data)
```
Based on the reference of the dataset, the following descriptions are given for 
the attributes in the dataset.

fixed acidity:

*   Most acids involved with wine or fixed or nonvolatile 
    (do not evaporate readily).

volatile acidity:

*   The amount of acetic acid in wine, which at too high of levels can lead to 
    an unpleasant, vinegar taste

citric acid:

*   Found in small quantities, citric acid can add 'freshness' and flavor 
    to wines

residual sugar:

*   The amount of sugar remaining after fermentation stops, it's rare to 
    find wines with less than 1 gram/liter and wines with greater than 
    45 grams/liter are considered sweet

chlorides: 

*   The amount of salt in the wine

free sulfur dioxide:

*   The free form of SO2 exists in equilibrium between molecular SO2 
    (as a dissolved gas) and bisulfite ion; it prevents microbial 
    growth and the oxidation of wine

total sulfur dioxide:

*   Amount of free and bound forms of S02; in low concentrations, SO2 is mostly 
    undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 
    becomes evident in the nose and taste of wine

density: 

*   The density of water is close to that of water depending on the percent 
    alcohol and sugar content

pH: 
*   Describes how acidic or basic a wine is on a scale from 0 (very acidic) to 
    14 (very basic); most wines are between 3-4 on the pH scale

sulphates:

*   A wine additive which can contribute to sulfur dioxide gas (S02) levels, 
    wich acts as an antimicrobial and antioxidant

alcohol: 

*   The percent alcohol content of the wine

quality: 

*   Score between 0 and 10)



The following snippet of data shows the first six records of the dataset.

```{r, echo=FALSE}
head(win_data)
```

A summary of the main attributes are given below. 

```{r, echo=FALSE}
summary(win_data[2:12])

```

### Univariate Plots Section

```{r, echo=FALSE, fig.height= 7}

## Making histograms and barcharts for the main atributes
p1 = ggplot(aes(x = fixed.acidity), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .25, color = 'black', 
                       fill = '#F79420')

p2 = ggplot(aes(x = volatile.acidity), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420') 

p3 = ggplot(aes(x = citric.acid), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420') 

p4 = ggplot(aes(x = residual.sugar), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 1, color = 'black', 
                       fill = '#F79420')

p5 = ggplot(aes(x = chlorides), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .007, color = 'black', 
                       fill = '#F79420') 

p6 = ggplot(aes(x = free.sulfur.dioxide), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 6, color = 'black', 
                       fill = '#F79420')

p7 =  ggplot(aes(x = total.sulfur.dioxide), data = win_data) + 
        geom_histogram(bins = 35, binwidth = 8, color = 'black', 
                       fill = '#F79420')

p8 = ggplot(aes(x = density), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .001, color = 'black', 
                       fill = '#F79420')

p9 =  ggplot(aes(x = pH), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .05, color = 'black', 
                       fill = '#F79420')

p10 = ggplot(aes(x = sulphates), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .02, color = 'black', 
                       fill = '#F79420')

p11 = ggplot(aes(x = alcohol), data = win_data) + 
        geom_histogram(bins = 35, binwidth = .25, color = 'black', 
                       fill = '#F79420') 

p12 = ggplot(aes(x = quality), data = win_data) + 
        geom_bar(color = 'black', fill = '#F79420') 
    

grid.arrange(p1, p2, p3,p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol = 3)

```

Based on the above grid of histograms and barchart, it is obvious that all the 
measured parameters for white wine are more or less normally distributed except 
'residual sugar' which is left skewed.

```{r, echo=FALSE}

```


```{r, echo=FALSE}

```



```{r, echo=FALSE}

```



```{r, echo=FALSE}

```






> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}

```

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!